---
title: "Statistics Group Project Using R"
author: "Libby Fortin (efortin@nd.edu), Keith O'Connor (koconn23@nd.edu), Chelsea Weibel (cweibel@nd.edu)"
date: "11/27/2017"
output: html_document
---

### Question 1: Evaluating the effect of three different new antibiotics on growth of E. coli in lab cultures.

```{r}
#### Linear plot of antibiotics effects on growth of E. coli ####

library(ggplot2)

antibiotics <- read.csv("antibiotics.csv")
      
plot1 <- ggplot(antibiotics, aes(x = trt, y = growth)) +
      geom_boxplot() + 
      theme_bw() + 
      xlab("Antibiotic Type") + ylab("E. coli Growth") +
      theme(plot.title = element_text(hjust = 0.5)) + # Centers the title
      ggtitle("Different Antibiotics on Growth")
                        
plot1

```

#### Next, we need to look at antibiotic treatments on the growth of E. coli using an ANOVA-design linear model and likelihood ratio test. (11/29/17)


```{r}

```

### Question 2: Evaluating the effect of sugar concentration on growth of E. coli in lab cultures.
```{r}
#### Linear plot of sugar effects on growth with line equation and R2 ####

sugardata <- read.csv(file = "sugar.csv", header=TRUE)

plot2 <- ggplot(data = sugardata, aes(x = sugar, y = growth))+
  geom_point(color = "blue1", shape = 16, size = 2) +
  theme_classic() +
  xlab("sugar") + ylab("growth") +
  ggtitle("Sugar vs. Growth") +
  theme(plot.title = element_text(hjust = 0.5)) + #Centers the title
  stat_smooth(method = "lm", se=FALSE, color="black") + #Adds a trend line, se=FALSE gets rid of cloud around trendline.
  geom_text(aes(x = 5, y = 25, label = lm_eqn1(lm(sugar ~ growth, sugardata))), parse = TRUE) # Adds the values for the R2 value and equation of the trendline to the coordinates x and y

m= lm(sugar ~ growth, sugardata)

lm_eqn1 = function(m) {       #Function to calculate and add R2 value and equation of the trendline to plot
  
  l <- list(a = format(coef(m)[1], digits = 2),
            b = format(abs(coef(m)[2]), digits = 2),
            r2 = format(summary(m)$r.squared, digits = 3));
  
  if (coef(m)[2] >= 0)  {
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
  } else {
    eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l)    
  }
  
  as.character(as.expression(eq));                 
}

plot2

```

#### Next, we need test for an effect of sugar concentration on growth of E. coli using a regression-design linear model and likelihood ratio test. (11/29/17)
